---
title: "Stroop Effect Exploration By Mehdi Jafari"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages} 
library(ggplot2) 
library(gridExtra)

``` 

## Introduction

In psychology, the Stroop effect is a demonstration of interference in the 
reaction time of a task. When the name of a color (e.g., "blue", "green", or 
"red") is printed in a color that is not denoted by the name (e.g., the word 
"red" printed in blue ink instead of red ink), naming the color of the word 
takes longer and is more prone to errors than when the color of the ink matches
the name of the color.

In a Stroop task, participants are presented with a list of words, with each 
word displayed in a color of ink. The participant’s task is to say out loud the
color of the ink in which the word is printed. The task has two conditions: a 
congruent words condition, and an incongruent words condition. In the congruent
words condition, the words being displayed are color words whose names match the
colors in which they are printed: for example, <span style="color:red">RED
</span>, <span style="color:blue">BLUE</span> In the incongruent 
words condition, the words displayed are color words whose names do not match 
the colors in which they are printed: for example, <span style="color:PURPLE">
PURPLE</span>, <span style="color:ORANGE">ORANGE</span>. In each case,
we measure the time it takes to name the ink colors in equally-sized lists. 
Each participant will go through and record a time from each condition.


```{r echo=FALSE, message=FALSE, warning=FALSE} 
# Load the Data 
dataset <- read.csv('dataset.csv') 
```

## Questions

1- What is our independent variable? What is our dependent variable?

  + dependent variable is the time that takes each participant to name the ink 
    color
  + independent variable is the congruent or incongruent of shown words




2- What is an appropriate set of hypotheses for this task? What kind of 
statistical test do you expect to perform? Justify your choices.

Null hypotheses => There is no difference in population means of response time 
under incongruent and congruent scenarios, which can be mathematically 
represented as:

µi = µ0 

Alternative hypotheses => population means of response time under incongruent 
condition is greater than congruent condition's, which can be represented 
mathematically as:  

µi > µ0

Since the sample size is less than 30 and we don't have the standard deviation 
of the population then we should use t test statistics. If we had standard 
deviation of population and the sample size was greater than 30 we would go 
with z-test.

+ H0 --> µi = µ0
+ Ha --> µi > µ0

Confidence level is 95% then α is 0.05.

T critical value = 1.714

3- Report some descriptive statistics regarding this dataset. Include at least 
one measure of central tendency and at least one measure of variability

```{r echo=FALSE, message=FALSE, warning=FALSE}
#output summary of the data
summary(dataset)

sdCongurent <- sd(dataset$Congruent) 
sdIncongruent <- sd(dataset$Incongruent)

paste("Standard Deviation of Congurent variable", sd(dataset$Congruent), sep=" : ")
paste("Standard Deviation of Incongruent variable", sd(dataset$Incongruent), sep=" : ")

paste("Variance of Congurent variable", var(dataset$Congruent), sep=" : ")
paste("Variance of Incongruent variable", var(dataset$Incongruent), sep=" : ")

```

  
4- Provide one or two visualizations that show the distribution of the sample 
data. Write one or two sentences noting what you observe about the plot or plots
.

As it is depicted in the following histograms, most participants completed the
incongruent word set between 16-25 Seconds while the congruent word set took 
them between 11-17 seconds. It shows the congruent word set takes less time
than the incongruent word set.

Also by looking at the summary we can see the 75% of the distribution of 
congruent dataset is less than 16.20 seconds while in incongruent dataset 
is 24.5 seconds.


```{r echo=FALSE, message=FALSE, warning=FALSE}


p1 = ggplot( aes(x= Congruent), data=dataset) +
  geom_histogram(col="red", fill="green" , binwidth = 2)

p2 = ggplot( aes(x= Incongruent), data=dataset) +
  geom_histogram(col="red", fill="green" , binwidth = 2)
grid.arrange(p1,p2 , ncol = 2)

```

5- Now, perform the statistical test and report your results. What is your 
confidence level and your critical statistic value? Do you reject the null 
hypothesis or fail to reject it? Come to a conclusion in terms of the experiment
task. Did the results match up with your expectations?

Calculated values based on the dataset are the follwoing:  

+ Mean of differences	: 7.96  

+ Standard Deviation of differences :	4.86482691  

+ T-critical value : 1.74  

+ Degree of freedom	: 23  

+ Direction of the test : One tailed test  

+ <span>α</span>	: 0.05  

+ Standard Error (SEM) : 0.993028635  

+ T-Statistics : 7.964791667  

+ Cohen's D	: 1.637219949  

+ R2 : 0.733913037  

+ Margin of Error :	1.727869825  

+ Confidence Interval : 6.24 <=  CI  => 9.69  

This is a dependent-sample and a pre-test, post-test scenario.

Since the t-statistics falls in critical region which is way bigger than 
1.714(t-critical value for α=0.05) we reject the null hypotheses.

P-value is less than .05 Which means the two sample populations are 
statistically significant.

R2 is equal to 73% which means 73% of the difference is due to the incongruent
treatment.

From the experiment we saw that there is a significant difference between the 
mean response times for the congruent and incongruent tests.With the help of 
the t-test performed we can infer that there is a significant difference in 
the time taken for the two tests in the population. This result matched with
 my expectation based on the Stroop test I have taken.
